We model a population of sexual organisms evolving in a landscape with two habitat patches connected by dispersal. Two resources are available to the organisms to feed on, and reproductive success is determined by the attack rates of the organisms on the resources. Attack rates on the two resources are determined by an underlying quantitative ecological trait $x$, which is subject to a trade-off so as to favor ecological divergence into two resource specialist strategies. The trait is underlain by a single haploid locus with a continuum of alleles. The ecological trait is also subject to sexual selection in the form of assortative mating, where females tend to prefer mates with more similar trait values.

\subsection*{Life cycle}

The life cycle of the organisms unfolds in discrete generations. At each generation the population goes through a phase of competition for the limited resources, followed by reproduction, survival and dispersal. Population size is variable and generations overlap.

\paragraph{Competition} Individuals compete for the resources available in their local habitat-patch. The competitive ability of an individual depends on its trait value $x$ and determines its reproductive success, or fitness $W_j(x)$ in patch $j$, as follows:

\begin{equation}
    W_j(x) = e_1(x) \, R^*_{1j} + e_2(x) \, R^*_{2j}
    \label{W_j}
\end{equation}

where $e_i(x)$ is the attack rate of the individual on resource $i$ and $R^*_{ij}$ is the equilibrium resource concentration of resource $i$ in patch $j$. The attack rate of the individual on resource $i$ depend on the ecological trait $x$ as follows

\begin{equation}
    e_i(x) = \exp{(-s \, (x \pm 1)^2)}
    \label{eq:attack_rates}
\end{equation}

% Feeding efficiencies depend on the ecological trait
\begin{figure}
    \begin{center}
        \begin{tikzpicture}
            \begin{axis}[xmin=-3, xmax=3, ymin=0, ymax=1.1, width=7cm, height=5cm, xtick distance=1,ytick distance=0.2, xlabel=Ecological trait ($x$), ylabel=Attack rate, legend entries={Resource 1, Resource 2}, legend style={draw=none, legend pos=outer north east}]
                \addplot[orange, samples=1000] {exp(-2*(x-1)^2)};
                \addplot[green, samples=1000] {exp(-2*(x+1)^2)};
                \addplot[densely dotted, green, samples=100] {exp(-0.1*(x+1)^2)};
                \addplot[dashed, green, samples=100] {exp(-0.5*(x+1)^2)};
                \addplot[densely dotted, orange, samples=100] {exp(-0.1*(x-1)^2)};
                \addplot[dashed, orange, samples=100] {exp(-0.5*(x-1)^2)};
            \end{axis}
            \end{tikzpicture}
        \end{center}
    \label{fig:attack_rates}
    \caption{Attack rate on both resources as a function of the ecological trait. The ecological selection coefficient $s$ determines the strength of the trade-off: dotted lines, $s = 0.1$, dashed lines, $s = 0.5$, solid lines, $s = 2$.}
\end{figure}

where $s$ is the ecological divergent selection coefficient. The respective attack rates on the first and second resource are maximized at values of $x$ of $-1$ and $+1$, which represent specialist strategies. The attack rates are subject to a trade-off whose strength is controlled by $s$, such that organisms cannot have a high payoff on both resources. The stronger this ecological trade-off, the more generalist phenotypes, i.e. values of $x$ near zero, suffer reduced attack rates relative to specialist competitors (Fig. \ref{fig:attack_rates}). If $s = 0$ the ecological trait does not affect the attack rates, i.e. there is no selection.\\

The resources flow through the landscape like through a chemostat according to the following dynamics:

\begin{equation}
    \frac{dR_{ij}}{dt} = a \, H_{ij} - \big(b + \sum_{k \in S_j} e_i(x_k) \big) \, R_{ij}
    \label{eq:resource_dynamics}
\end{equation}

where $R_{ij}$ is the concentration of resource $i$ in habitat-patch $j$, $a$ is the absolute rate of resource inflow, $b$ is the relative rate of resource outflow, $H_{ij}$ is a coefficient of symmetry in rates of resource inflow between the two patches, and $e_i(x_k)$ is the attack rate of individual $k$ on resource $i$. The sum of attack rates across the set $S_j$ of individuals living in patch $j$ symbolizes the local consumption of the resource by the organisms. The habitat symmetry factor $H_{ij}$ controls the reduction in inflow of resource 2 relative to resource 1 in patch 1, and of resource 1 relative to resource 2 in patch 2, such that the two habitat patches are mirror-images of each other with respect to resource inflow,

\begin{equation}
    H = 
    \begin{pmatrix} 
        1 & h \\ 
        h & 1
    \end{pmatrix}
\end{equation}

where $h$ is the habitat symmetry parameter, ranging from 0 to 1. When $h = 1$, the rate of inflow is homogeneous between resources across habitats. On the contrary, when $h = 0$, the landscape is fully heterogeneous in resource inflow, where only resource 1 flows through patch 1 and only resource 2 flows through patch 2. The level of environmental heterogeneity is an important determinant of the mode of speciation that will occur (Rettelbach).\\

We assume that the dynamics of the resources are fast relative to the population dynamics, such that equilibrium resource concentrations are given by solving Equation \ref{eq:resource_dynamics}:

\begin{equation}
    R^*_{ij} = \frac{a \, H_{ij}}{b + \sum_{k \in S_j} e_i(x_k)}
\end{equation}

\paragraph{Reproduction} Females and males encounter each other and mate within patches. We assume a balanced sex ratio. The probability that female $\female$ accepts encountered male $\male$, i.e. the attractiveness of male $\male$ to female $\female$, depends on both their ecological trait values and is given by the assortative mating probability function

\begin{equation}
    A(x_{\male}, x_{\female}) = \exp{\big(-\alpha \, (x_{\male} - x_{\female})^2\big)}
    \label{eq:assortative_mating}
\end{equation}

where $\alpha$ is the sexual selection coefficient, scaling the sensitivity of females to deviations in ecological trait values. Females have a stronger preference for more ecologically similar males as $\alpha$ increases (Fig. \ref{fig:mating_probability}). When $\alpha = 0$, females mate at random.\\ 

\begin{figure}
    \begin{center}
        \begin{tikzpicture}
            \begin{axis}[xmin=-2, xmax=2, ymin=0, ymax=1.1, width=7cm, height=5cm, xtick distance=1,ytick distance=0.2, xlabel=Distance in ecological trait ($x_{\male} - x_{\female}$), ylabel=Mating probability, legend entries={$\alpha = 0.1$, $\alpha = 1$, $\alpha = 10$, $\alpha = 100$}, legend style={font=\large, draw=none, legend pos=outer north east}]
                \addplot[loosely dotted, red, samples=100] {exp(-0.1*x^2)};
                \addplot[densely dotted, red, samples=100] {exp(-1*x^2)};
                \addplot[dashed, red, samples=1000] {exp(-10*x^2)};
                \addplot[red, samples=1000] {exp(-100*x^2)};
            \end{axis}
            \end{tikzpicture}
        \end{center}
    \label{fig:mating_probability}
    \caption{Mating probability, or attractiveness, of a male with trait value $x_i$ to a female with trait value $x_j$ for different values of the sexual selection coefficient $\alpha$.}
\end{figure}

The competitive abilities $W_j$ affect the reproductive output of males and females differently. In females, $W_j(x_{\female})$ is the number of offspring that a female produces. In males, $W_j(x_{\male})$ is the rate of encounter by females, and is analogous to the activity of the male on the mating market.

\paragraph{Survival} Newborns of the current generation all survive to the next generation and become sexually mature. Generations overlap and adults of the current generation survive with probability $1-d$, where $d$ is the per capita death rate.

\paragraph{Dispersal} At very generation a proportion $m$ of the individuals disperses to the alternative patch.

\subsection*{Adaptive dynamics}

We use adaptive dynamics to analyze the evolutionary outcome of our model under the effects of natural and sexual selection. We first describe the dynamics of a mutant arising in a monomorphic population with respect to trait $x$. Taking the geometric growth rate of the mutant as a measure of invasion fitness, we then derive an expression for the selection gradient and the curvature of the fitness function, which can be used to predict the endpoints of the evolutionary dynamics and how they depend on model parameters.

\subsubsection*{Invasion fitness}

Adaptive dynamics theory is concerned with the conditions under which an initially rare mutant can spread in a population of resident phenotypes (cite Geritz and Metz). The derivations are simplified by assuming that the resident population is monomorphic with respect to the evolving trait, here $x$. In a model with discrete-time population dynamics, a mutant will spread if its geometric growth rate, or invasion fitness, is higher the growth rate of the resident. Assuming that (1) mutations are rare relative to the population dynamics, (2) selection is the only force at play (no drift) and (3) the mutant allele is only slightly deviating from the resident allele in phenotype, a mutant that spreads will invade the population and become the new resident (cite?).\\

We can derive a recursion equation for the population dynamics of a monomorphic resident population of our study organism based on the aforementioned life cycle:

\begin{equation}
    \overrightarrow{N'}_{t+1} = \pmb{\Lambda} \, \overrightarrow{N'}_t
\end{equation}

where $\overrightarrow{N'}_t$ is the vector of mutant population densities across patches, $N'_1$ and $N'_2$, at discrete generation $t$. The transition matrix $\pmb{\Lambda}$ captures the steps of the life cycle of the organism,

\begin{equation}
    \pmb{\Lambda} = \pmb{M} \, \pmb{Q}
\end{equation}

where $\pmb{M}$ is the migration matrix, given by

\begin{equation}
    \pmb{M} =
    \begin{pmatrix}
        1-m & m \\
        m & 1-m 
    \end{pmatrix}
\end{equation}

and $\pmb{Q}$ is the matrix of local geometric per capita growth rates,

\begin{equation}
    \pmb{Q} =
    \begin{pmatrix}
        r_1(x, \hat{x}) & 0 \\
        0 & r_2(x, \hat{x})
    \end{pmatrix}
\end{equation}

The within-patch geometric growth rate captures local competition, reproduction and survival. The local growth rate in patch $j$ is given by

\begin{equation}
    r_j(x, \hat{x}) = 1 - d + \frac{1}{2} \, \Bigg[ W_j(x, \hat{x}) + W_j(\hat{x}, \hat{x}) \, \frac{W_j(x, \hat{x})\,A(x,\hat{x})}{W_j(\hat{x}, \hat{x})\,A(\hat{x},\hat{x})}\Bigg]
    \label{eq:growth_rate}
\end{equation}

where $W_j(x,\hat{x})$ is the reproductive success in patch $j$ of a mutant with trait value $x$ in a resident population with trait value $\hat{x}$, and $A(x, \hat{x})$ is the attractiveness of a male mutant to a resident female. The factor between the square brackets is the number of offspring produced by the mutant population. The first term within this factor is the number of offspring directly produced by mutant females, $W_j(x, \hat{x})$. The second term is the number of offspring contributed for by males, who do not produce offspring directly but contribute to the number of offspring their mated partner produces. Here, we assume that the mutant is sufficiently rare so that mutant males will only mate with resident females and thus contribute to $W_j(\hat x, \hat x)$ offspring. This reproductive output is weighed by the mating success of the males, which is their reproductive success on the male market $W(x, \hat{x})$, times their attractiveness to a resident female $A(x, \hat{x})$, relative to the mating success of the resident males, $W_j(\hat x, \hat x)\,A(\hat x, \hat x)$. Finally, in a haploid system with Mendelian inheritance, only half of the offspring of mutants mating with residents will carry the mutant allele.\\

Here, we assume that the mutant is rare enough so that only the resident affects the resource dynamics. The reproductive success of a mutant is therefore

\begin{equation}
    W_j(x, \hat x) = e_1(x)\,R_{1j}^*(\hat x) + e_2(x)\,R_{2j}^*(\hat x)
\end{equation}

where the equilibrium resource concentrations depend on the resident trait value,

\begin{equation}
    R_{ij}^*(\hat x) = \frac{a\,H_{ij}}{b + N_j^*(\hat x) \, e_i(\hat x)}
\end{equation}

where $N_j^*(\hat x)$ is the equilibrium population density of the resident in patch $j$.\\

Given that $A(\hat{x}, \hat{x}) = 1$, Equation \ref{eq:growth_rate} simplifies to

\begin{equation}
    r_j(x, \hat{x}) = 1 - d + \frac{1}{2} \, W(x, \hat{x}) \, \big(1 + A(x, \hat{x})\big)
\end{equation}

The invasion fitness, i.e. the long-term growth rate of a mutant arising in a resident population at its demographic equilibrium, is the leading eigenvalue of the transition matrix $\pmb{\Lambda}$,

\begin{equation}
    \lambda(x, \hat{x}) = \frac{1}{2} \bigg[(1-m) \, (r_1 + r_2) + \sqrt{(1-m)^2 \, (r_1^2 - 2 \, r_1 \, r_2 + r_2^2) + 4 \, m^2 \, r_1 \, r_2} \bigg]
    \label{eq:invasion_fitness}
\end{equation}

where $r_j = r_j(x, \hat x)$. A mutant with trait $x$ can invade a resident population with trait $\hat x$ if its invasion fitness is greater than one. Next, we use the invasion fitness to derive the direction of evolution by selection.

\subsubsection*{Selection gradient}

If evolution is limited by mutations and occurs through successive allele substitutions, the direction and extent of evolution by selection is described by the selection gradient, which is the change in invasion fitness as the trait value of the mutant changes, given a resident trait value:

\begin{equation}
    G(\hat{x}) = \frac{\partial \lambda}{\partial x}\bigg|_{x=\hat{x}}
\end{equation}

A positive selection gradient will favor an increase in trait value while a negative selection gradient will favor a decrease. This could eventually lead to an evolutionary steady-state, but could also trigger a runaway to infinite values of trait $x$.\\

The derivative of the invasion fitness can be written as:

\begin{equation}
   \frac{\partial \lambda}{\partial x}\bigg = \frac{1}{\overrightarrow{v}^T\,\overrightarrow{u}} \, \bigg( \overrightarrow{v}^T \, \frac{\partial \pmb{\Lambda}}{\partial x} \, \overrightarrow{u} \bigg)
   \label{eq:deriv_fitness}
\end{equation}

(Otto and Day) where $\overrightarrow{v}$ and $\overrightarrow{u}$ are left and right dominant eigenvectors, respectively, of the transition matrix $\pmb{\Lambda}$. Possible such eigenvectors are

\begin{equation}
    \overrightarrow{v} = 
    \begin{pmatrix}
        \lambda - (1-m)\,r_2 \\
        m\,r_2 
    \end{pmatrix}
    \label{eq:left_eigenvector}
\end{equation}

and

\begin{equation}
    \overrightarrow{u} = 
    \begin{pmatrix}
        \lambda - (1-m)\,r_2 \\
        m\,r_1
    \end{pmatrix}
\end{equation}

where $\lambda = \lambda(x, \hat x)$. The derivative of the transition matrix $\pmb\Lambda$ with respect to $x$ is

\begin{equation}
    \frac{\partial \pmb{\Lambda}}{\partial x} = \pmb{M} \, \frac{\partial \pmb{Q}}{\partial x}
\end{equation}

where $\partial \pmb{Q} / \partial x$ is a diagonal matrix whose elements are, the derivatives of the local geometric growth rates in each patch $j$,

\begin{equation}
    \frac{\partial r_j}{\partial x} = \frac{1}{2} \, \bigg[\frac{\partial W_j}{\partial x} \, (1 + A(x, \hat{x})) + \frac{\partial A}{\partial x} \, W_j(x, \hat{x})\bigg]
    \label{eq:deriv_local_growth}
\end{equation}

where the derivative of the attractiveness function is

\begin{equation}
    \frac{\partial A}{\partial x} = -2 \, \alpha \, (x - \hat{x}) \, A(x, \hat{x})
\end{equation}

At the resident trait value, $\partial A / \partial x = 0$ and $A(\hat{x},\hat{x}) = 1$, and so the derivative of the local growth rate reduces to

\begin{equation}
    \frac{\partial r_j}{\partial x}\bigg|_{x=\hat{x}} = \frac{\partial W_j}{\partial x}\bigg|_{x=\hat{x}} 
\end{equation}

We also know that $\lambda(\hat{x}, \hat{x}) = 1$, so the eigenvectors $\overrightarrow{v}$ and $\overrightarrow{u}$ evaluated at the resident simplify to

\begin{equation}
    \hat{\overrightarrow{v}} = 
    \begin{pmatrix}
        1 - (1-m)\,\hat{r}_2\\
        m \, \hat{r}_2
    \end{pmatrix}
    \label{eq:left_eigenvector_resident}
\end{equation}

and

\begin{equation}
    \hat{\overrightarrow{u}} = 
    \begin{pmatrix}
        1 - (1-m)\,\hat{r}_2\\
        m \, \hat{r}_1
    \end{pmatrix}
    \label{eq:right_eigenvector_resident}
\end{equation}

where the local growth rate of the resident is

\begin{equation}
    \hat{r}_j = r_j(\hat{x}, \hat{x}) = 1 - d + W_j(\hat{x}, \hat{x})
    \label{eq:local_growth_resident}
\end{equation}

Substituting Equations \ref{eq:left_eigenvector}--\ref{eq:local_growth_resident} into Equation \ref{eq:deriv_fitness} and simplifying, we get the following expression for the selection gradient:

\begin{equation}
    G(\hat{x}) = \frac{\big[1-m-(2-4\,m+m^2)\,\hat{r}_2+(1-3\,m+2\,m^2)\,\hat{r}^2_2\big]\, \hat{W}_1' + m^2\,\hat{r}_1\,\hat{W}_2'}{(1-\hat{r}_2+m\,\hat{r}_2)^2 + m^2\,\hat{r}_1\,\hat{r}_2}
    \label{eq:fitness_gradient_expression}
\end{equation}

where $\hat{W}_j' = \partial W_j / \partial x |_{x = \hat{x}}$. The selection gradient does not involve the sexual selection coefficient $\alpha$, hence, the values of the singular strategies depend only on competition, survival and migration, but not on sexual selection. Next, we show that sexual selection nevertheless affects the evolutionary stability of the singular strategies.

\subsubsection*{Evolutionary stability}

Evolutionary singular strategies are trait values for which the selection gradient is zero, and correspond to either local minima or local maxima in the invasion fitness landscape. A singular strategy that is a local maximum is invasion-proof and is evolutionarily stable. A local minimum is a resident trait value that can be invaded by any neighboring mutant and is evolutionarily unstable, i.e. it is under disruptive selection and promotes the branching and coexistence of multiple trait values. The evolutionary stability of a singular strategy is given by the curvature of the invasion fitness function at the singular trait value,

\begin{equation}
    \frac{\partial^2 \lambda}{\partial x^2}\bigg|_{x=\hat{x}=x^*}
\end{equation}

The curvature of the invasion fitness function can be obtained by differentiating its derivative,

\begin{equation}
    \frac{\partial^2 \lambda}{\partial x^2} = \frac{\overrightarrow{v}^T\,\overrightarrow{u}\,\frac{\partial}{\partial x}\,\big(\overrightarrow{v}^T\,\frac{\partial \pmb{\Lambda}}{\partial x}\,\overrightarrow{u}\big) - \frac{\partial}{\partial x} \big( \overrightarrow{v}^T \, \overrightarrow{u} \big) \, \overrightarrow{v}^T \, \frac{\partial \pmb{\Lambda}}{\partial x}\,\overrightarrow{u}}{(\overrightarrow{v}^T\,\overrightarrow{u})^2}
\end{equation}

The selection gradient is zero at the singular strategy, or equilibrium, $x^*$, i.e.

\begin{equation}
    \bigg( \overrightarrow{v}^T\,\frac{
    \partial \pmb \Lambda}{\partial x}\,\overrightarrow{u} \bigg)_* = G(x^*) = 0
\end{equation}

Therefore, the curvature of the fitness function at equilibrium $x^*$, simplifies to

\begin{equation}
    \frac{\partial^2 \lambda}{\partial x^2}\bigg|_* = \Bigg[ \frac{1}{\overrightarrow{v}^T\,\overrightarrow{u}}\,\frac{\partial}{\partial x}\,\bigg(\overrightarrow{v}^T\,\frac{\partial \pmb{\Lambda}}{\partial x}\,\overrightarrow{u}\bigg) \Bigg]_*
    \label{eq:fitness_curvature_equilibrium}
\end{equation}

where

\begin{equation}
    \frac{\partial}{\partial x} \, \bigg(\overrightarrow{v}^T \, \frac{\partial \pmb{\Lambda}}{\partial x} \, \overrightarrow{u}\bigg) = \frac{\partial \overrightarrow{v}^T}{\partial x}\,\frac{\partial \pmb{\Lambda}}{\partial x}\,\overrightarrow{u} + \overrightarrow{v}^T\,\frac{\partial^2 \pmb{\Lambda}}{\partial x^2}\,\overrightarrow{u} + \overrightarrow{v}^T\,\frac{\partial \pmb{\Lambda}}{\partial x}\,\frac{\partial \overrightarrow{u}}{\partial x}
    \label{eq:deriv_num_gradient_unevaluated}
\end{equation}

where the derivatives of the left and right eigenvectors are

\begin{equation}
    \frac{\partial \overrightarrow{v}}{\partial x} = 
    \begin{pmatrix}
        \nicefrac{\partial \lambda}{\partial x} - (1-m)\,\nicefrac{\partial r_2}{\partial x}\\
        m\,\nicefrac{\partial r_2}{\partial x}
    \end{pmatrix}
\end{equation}

and

\begin{equation}
    \frac{\partial \overrightarrow{u}}{\partial x} = 
    \begin{pmatrix}
        \nicefrac{\partial \lambda}{\partial x} - (1-m)\,\nicefrac{\partial r_2}{\partial x}\\
        m\,\nicefrac{\partial r_1}{\partial x}
    \end{pmatrix}
\end{equation}

which, because $\partial r_j / \partial x = \partial W_j / \partial x$ at the resident $\hat x$ and $\partial \lambda / \partial x = 0$ at the equilibrium $x^*$, respectively evaluate to

\begin{equation}
    \frac{\partial \overrightarrow{v}}{\partial x}\bigg|_{x=\hat x=x^*} = 
    \begin{pmatrix}
        (m-1)\,\nicefrac{\partial W_2}{\partial x}\\
        m\,\nicefrac{\partial W_2}{\partial x}
    \end{pmatrix}
    _{x=\hat x=x^*}
\end{equation}

and

\begin{equation}
    \frac{\partial \overrightarrow{u}}{\partial x}\bigg|_{x=\hat x=x^*} = 
    \begin{pmatrix}
        (m-1)\,\nicefrac{\partial W_2}{\partial x}\\
        m\,\nicefrac{\partial W_1}{\partial x}
    \end{pmatrix}
    _{x=\hat x=x^*}
    \label{eq:right_eigenvector_equilibrium}
\end{equation}

The second derivative of the transition matrix $\pmb \Lambda$ is given by 

\begin{equation}
    \frac{\partial^2 \pmb{\Lambda}}{\partial x^2} = \pmb{M} \, \frac{\partial^2 \pmb{Q}}{\partial x^2}
    \label{eq:second_derivative_transition}
\end{equation}

where $\partial^2 \pmb{Q} / \partial x^2$ is a diagonal matrix whose elements are the second derivatives of the local growth rates for each patch $j$, obtained by differentiating their first-order derivatives in Equation \ref{eq:deriv_local_growth}:

\begin{equation}
    \frac{\partial^2 r_j}{\partial x^2} = \frac{1}{2} \bigg[\frac{\partial^2 W_j}{\partial x^2}\,\big(1+A(x,\hat{x})\big) + 2\,\frac{\partial W_j}{\partial x}\,\frac{\partial A}{\partial x} + \frac{\partial^2 A}{\partial x^2}\,W_j(x,\hat{x})\bigg]
\end{equation}

where the second derivative of the attractiveness function is

\begin{equation}
    \frac{\partial^2 A}{\partial x^2} = -2\,\alpha\,\big( A(x,\hat{x}) + (x-\hat{x})\,\nicefrac{\partial A}{\partial x} \big)
\end{equation}

which, at equilibrium $x=\hat{x}=x^*$ becomes

\begin{equation}
    \frac{\partial^2 A}{\partial x^2}\bigg|_{x=\hat{x}=x^*} = -2\,\alpha
\end{equation}

The second derivatives of the local growth rates evaluated at equilibrium $x^*$ therefore simplify to

\begin{equation}
    \frac{\partial^2 r_j}{\partial x^2}\bigg|_{x=\hat{x}=x^*}= \frac{\partial^2 W_j}{\partial x^2}\bigg|_{x=\hat{x}=x^*} - \alpha \, W_j(x^*,x^*)
    \label{eq:second_derivative_local_growth_equilibrium}
\end{equation}

By substituting Equations \ref{eq:left_eigenvector_resident}, \ref{eq:right_eigenvector_resident} and \ref{eq:deriv_num_gradient_unevaluated}--\ref{eq:second_derivative_local_growth_equilibrium} into Equation \ref{eq:fitness_curvature_equilibrium} and simplifying, we obtain an expression for the curvature of the fitness function at equilibrium:

\begin{equation}
    \begin{split}
        \frac{\partial^2 \lambda}{\partial x^2}\bigg|_{x=\hat x=x^*}=&\frac{1}{1 + m^2\,r^*_1\,r^*_2 + (m - 1)\,\big(2 + (m - 1)\,r^*_2^2\big)} \times \\
        &\Bigg[m^2\, r^*_ 1\, W''^*_2 + 2\, (2\, m - 1)\, \big(1 + (m - 1)\, r^*_ 2\big)\, W'^*_1 \, W'^*_2 +\\
        &\big(1 + (m - 1) \, r^*_ 2\big)\, \big(1 - m + (2 m - 1) \,r^*_ 2\big)\, W''^*_1 -\\
        &\alpha \times \Big(m^2\, r^*_ 1\, W^*_2 - \\
        & \big(m - 1 + (2 - 4\,m + m^2)\, r^*_ 2 - (1 - 3\,m + 2\,m^2)\,r^*_ 2^2\big)\,W^*_1 \Big)\Bigg]
    \end{split}
    \label{eq:fitness_curvature_equilibrium_expression}
\end{equation}

where $W_j^* = W_j(x^*, x^*)$, $W_j''^* = \partial^2 W_j / \partial x^2 |_{x=\hat{x}=x^*}$ and $r_j^* = r_j(x^*, x^*)$. We used this expression to assess the stability of evolutionary equilibria. We now show that increasing the sexual selection coefficient $\alpha$, i.e. stronger assortative mating, has a stabilizing effect on the singular strategy and may counteract diversification.

\subsubsection*{The stabilizing effect of assortative mating}

Assume that the local per capita growth rate $r_j$ depends on two traits: $x$, which affects competitive ability, and $y$, which affects attractiveness. Those two traits are eventually the same, but we split them to study the effect of the ecological trait on ecological and sexual selection separately. We first show that $x$ and $y$ have independent effects on the curvature of the fitness function at equilibrium, that is,

\begin{equation}
    \frac{\partial^2 \lambda}{\partial x \partial y}\bigg|_* = 0
\end{equation}

where $*$ symbolizes $x=\hat x=x^*$, $y=\hat y=y^*$. The cross-derivative of the fitness function with respect to $x$ and $y$ is

\begin{equation}
    \frac{\partial^2 \lambda}{\partial x \partial y} = \frac{\partial}{\partial y}\,\Bigg[\frac{1}{\overrightarrow{v}^T\,\overrightarrow{u}}\,\bigg( \overrightarrow{v}^T\,\frac{\partial \Lambda}{\partial x}\,\overrightarrow{u} \bigg)\Bigg]
\end{equation}

which is equivalent to

\begin{equation}
    \frac{\partial^2 \lambda}{\partial x \partial y} = \frac{\frac{\partial}{\partial y}\Big(\overrightarrow{v}^T\,\frac{\partial \pmb \Lambda}{\partial x}\,\overrightarrow{u}\Big)\,\overrightarrow{v}^T\,\overrightarrow{u}-\frac{\partial}{\partial y}\,\Big(\overrightarrow{v}^T\,\overrightarrow{u}\Big)\,\Big(\overrightarrow{v}^T\,\frac{\partial \pmb \Lambda}{\partial x}\,\overrightarrow{u}\Big)}{(\overrightarrow{v}^T\,\overrightarrow{u})^2}
\end{equation}

in which

\begin{equation}
    \bigg(\overrightarrow{v}^T\,\frac{\partial \pmb \Lambda}{\partial x}\,\overrightarrow{u}\bigg)_* = 0
\end{equation}

as it corresponds to the fitness gradient, and

\begin{equation}
    \frac{\partial}{\partial y}\,\bigg(\overrightarrow{v}^T\,\frac{\partial \pmb \Lambda}{\partial x}\,\overrightarrow{u}\bigg) = \frac{\partial \overrightarrow{v}^T}{\partial y}\,\frac{\partial \pmb \Lambda}{\partial x}\,\overrightarrow{u}+\overrightarrow{v}^T\,\frac{\partial^2 \pmb \Lambda}{\partial x \partial y}\,\overrightarrow{u}+\overrightarrow{v}^T\,\frac{\partial \pmb \Lambda}{\partial x}\,\frac{\partial \overrightarrow{u}}{\partial y}   
    \label{eq:deriv_lambda_x_y}
\end{equation}

where the derivatives of the left and right eigenvectors with respect to $y$ are, respectively,

\begin{equation}
    \frac{\partial \overrightarrow{v}}{\partial y} =
    \begin{pmatrix}
        \nicefrac{\partial \lambda}{\partial y} - (1-m)\,\nicefrac{\partial r_2}{\partial y}\\
        m \,\nicefrac{\partial r_2}{\partial y}
    \end{pmatrix}
\end{equation}

and

\begin{equation}
    \frac{\partial \overrightarrow{u}}{\partial y} =
    \begin{pmatrix}
        \nicefrac{\partial \lambda}{\partial y} - (1-m)\,\nicefrac{\partial r_2}{\partial y}\\
        m \,\nicefrac{\partial r_1}{\partial y}
    \end{pmatrix}
\end{equation}

where $\partial \lambda / \partial y = 0$ at equilibrium $x^*$, $y^*$, and

\begin{equation}
    \frac{\partial r_j}{\partial y}\bigg|_* = \frac{1}{2} \, \frac{\partial A}{\partial y}\bigg|_*\,W_j(x^*,x^*)
\end{equation}

where $\partial A / \partial y = 0$ at equilibrium $y^*$. Therefore, both $\partial \overrightarrow{v} / \partial y$ and $\partial \overrightarrow{u} / \partial y$ are vectors of zeros at equilibrium. The cross-derivative of the transition matrix is

\begin{equation}
    \frac{\partial^2 \pmb \Lambda}{\partial x \partial y} = \pmb M \, \frac{\partial}{\partial y}\,\frac{\partial \pmb Q}{\partial x}
\end{equation}

where 

\begin{equation}
    \frac{\partial \pmb Q}{\partial x} = \frac{1}{2}\,\big(1+A(y,\hat y)\big)\,\pmb M \frac{\partial \pmb W}{\partial x}
\end{equation}

where $\pmb W$ is independent of trait $y$ and so

\begin{equation}
    \frac{\partial}{\partial y}\,\frac{\partial \pmb Q}{\partial x} = \frac{1}{2}\,\frac{\partial A}{\partial y}\,\pmb M\,\frac{\partial \pmb W}{\partial x}
\end{equation}

which also reduces to zero at equilibrium and therefore, 

\begin{equation}
    \frac{\partial^2 \lambda}{\partial x \partial y}\bigg|_* = 0
\end{equation}

Therefore, competition and mate choice have independent effects on the curvature of the fitness function, i.e.e the stability of equilibria. The effect of mate choice on the stability of an equilibrium is given by

\begin{equation}
    \frac{\partial^2 \lambda}{\partial y^2}\bigg|_*
\end{equation}

If this expression is negative, sexual selection is stabilizing the evolutionary dynamics of the ecological trait and may prevent branching. The curvature of the fitness function with respect to $y$ is given by

\begin{equation}
    \frac{\partial^2 \lambda}{\partial y^2} = \frac{1}{\overrightarrow{v}^T\,\overrightarrow{u}}\,\bigg(\overrightarrow{v}^T\frac{\partial \pmb \Lambda}{\partial y}\,\overrightarrow{u}\bigg)
\end{equation}

where 

\begin{equation}
    \frac{\partial \pmb \Lambda}{\partial y} = \pmb M \frac{\partial \pmb Q}{\partial y}
\end{equation}

where

\begin{equation}
    \frac{\partial \pmb Q}{\partial y} = \frac{1}{2}\,\frac{\partial A}{\partial y}\,W(x,\hat x)
\end{equation}

therefore reducing to

\begin{equation}
    \frac{\partial^2 \lambda}{\partial y^2} = \frac{1}{2}\,\frac{\partial}{\partial y}\,\bigg(\frac{\partial A}{\partial y}\, \frac{\overrightarrow{v}^T\,\pmb W\,\overrightarrow{u}}{\overrightarrow{v}^T\,\overrightarrow{u}}\bigg)
\end{equation}

which is equivalent to

\begin{equation}
    \frac{\partial^2 \lambda}{\partial y^2} = \frac{1}{2}\,\frac{\frac{\partial}{\partial y}\,\Big(\frac{\partial A}{\partial y}\,\overrightarrow{v}^T\,\pmb W\,\overrightarrow{u}\Big)\,\overrightarrow{v}^T\,\overrightarrow{u}-\frac{\partial}{\partial y}\Big(\overrightarrow{v}^T\,\overrightarrow{u}\Big)\frac{\partial A}{\partial y}\,\Big(\overrightarrow{v}^T\,\pmb W\,\overrightarrow{u}\Big)}{(\overrightarrow{v}^T\,\overrightarrow{u})^2}
\end{equation}

where $\partial A / \partial y = 0$ at equilibrium and which therefore simplifies to

\begin{equation}
    \frac{\partial^2 \lambda}{\partial y^2}\bigg|_* = \frac{1}{2}\,\Bigg(\frac{1}{\overrightarrow{v}^T\,\overrightarrow{u}}\,\frac{\partial}{\partial y}\,\bigg[\frac{\partial^2 A}{\partial y^2}\,\Big(\overrightarrow{v}^T\,\pmb W\,\overrightarrow{u}\Big)+\frac{\partial A}{\partial y}\,\frac{\partial \big(\overrightarrow{v}^T\,\pmb W\,\overrightarrow{u}\big)}{\partial y}\bigg]\Bigg)_*
\end{equation}

which further simplifies to

\begin{equation}
    \frac{\partial^2 \lambda}{\partial y^2}\bigg|_* = \frac{1}{2}\,\Bigg(\frac{\overrightarrow{v}^T\,\pmb W\,\overrightarrow{u}}{\overrightarrow{v}^T\,\overrightarrow{u}}\,\frac{\partial^2 A}{\partial y}\Bigg)_* = -\alpha \, \Bigg(\frac{\overrightarrow{v}^T\,\pmb W\,\overrightarrow{u}}{\overrightarrow{v}^T\,\overrightarrow{u}}\Bigg)_*
    \label{eq:curvature_final}
\end{equation}

where 

\begin{equation}
    \overrightarrow{v}^*^T\,\overrightarrow{u}^* = \big(1-(1-m)\,r^*_2\big)^2 + m^2\,r^*_1\,r^*_2 > 0
\end{equation}

assuming that the per capita growth rates $r^*_j$ have the same sign. The sign of the second derivative of $\lambda$ with respect to $y$, evaluated at the equilibrium, is therefore given by the numerator

\begin{equation}
    \overrightarrow{v}^*^T\,\pmb W^*\,\overrightarrow{u}^* = \big(1-(1-m)\,r^*_2\big)^2\,W^*_1 + m^2\,r^*_1\,r^*_2\,W^*_2
\end{equation}

which is also positive. The curvature of the fitness function $\partial^2 \lambda / \partial y^2$ at equilibrium thus depends negatively on the sexual selection coefficient $\alpha$, indicating that assortative mating stabilizes the evolutionary equilibrium.

\subsubsection*{Convergence stability}

Not every singular strategy is attainable through evolution. A singular strategy is convergence-stable, or attainable, if it is an attractor for nearby resident trait values, i.e. if the selection gradient pushes for an increase in trait value when slightly below the singular strategy, and for a decrease when slightly above. That is, 

\begin{equation}
    \frac{\partial G}{\partial \hat{x}}\bigg|_{\hat{x}=x^*} < 0
\end{equation}

where 

\begin{equation}
    \frac{\partial G}{\partial \hat{x}} = \frac{\partial}{\partial \hat{x}} \Bigg[ \frac{1}{\hat{\overrightarrow{v}}^T\,\hat{\overrightarrow{u}}} \bigg( \hat{\overrightarrow{v}}^T\,\frac{\partial \pmb{\Lambda}}{\partial x}\bigg|_{x=\hat{x}}\,\hat{\overrightarrow{u}} \bigg) \Bigg]
\end{equation}

which is equivalent to

\begin{equation}
    \frac{\partial G}{\partial \hat{x}} = \frac{\frac{\partial}{\partial \hat{x}} \big( \hat{\overrightarrow{v}}^T\,\frac{\partial \pmb{\Lambda}}{\partial x}\big|_{x=\hat{x}}\,\hat{\overrightarrow{u}} \big)\,\hat{\overrightarrow{v}}^T\,\hat{\overrightarrow{u}} - \big( \hat{\overrightarrow{v}}^T\,\frac{\partial \pmb{\Lambda}}{\partial x}\big|_{x=\hat{x}}\,\hat{\overrightarrow{u}} \big)\,\frac{\partial}{\partial \hat{x}} \big(\hat{\overrightarrow{v}}^T\,\hat{\overrightarrow{u}}\big)}{(\hat{\overrightarrow{v}}^T\,\hat{\overrightarrow{u}})^2}
    \label{eq:deriv_gradient}
\end{equation}

which, because

\begin{equation}
    \bigg( \overrightarrow{v}^T\,\frac{\partial \pmb \Lambda}{\partial x}\,\overrightarrow{u} \bigg)_* = G(x^*) = 0
    \label{eq:gradient_is_zero_at_equilibrium}
\end{equation}

simplifies, at equilibrium $x^*$, to

\begin{equation}
    \frac{\partial G}{\partial \hat x}\bigg|_* = \Bigg[ \frac{1}{\hat{\overrightarrow{v}}\,\hat{\overrightarrow{u}}} \, \frac{\partial}{\partial \hat x}\,\bigg(\hat{\overrightarrow{v}}^T\,\frac{\partial \pmb \Lambda}{\partial x}\bigg|_{x=\hat x}\,\hat{\overrightarrow{u}}\bigg)\Bigg]_*
    \label{eq:deriv_gradient_equilibrium}
\end{equation}

where

\begin{equation}
    \frac{\partial}{\partial \hat{x}} \bigg( \hat{\overrightarrow{v}}^T\,\frac{\partial \pmb{\Lambda}}{\partial x}\bigg|_{x=\hat{x}}\,\hat{\overrightarrow{u}} \bigg) = \frac{\partial \hat{\overrightarrow{v}}^T}{\partial \hat{x}}\,\frac{\partial \pmb{\Lambda}}{\partial x}\bigg|_{x=\hat{x}}\,\hat{\overrightarrow{u}} + \hat{\overrightarrow{v}}^T\,\frac{\partial}{\partial \hat{x}} \bigg(\frac{\partial \pmb{\Lambda}}{\partial x}\bigg|_{x=\hat{x}}\bigg)\,\hat{\overrightarrow{u}} + \hat{\overrightarrow{v}}^T\,\frac{\partial \pmb{\Lambda}}{\partial x}\bigg|_{x=\hat{x}}\,\frac{\partial \hat{\overrightarrow{u}}}{\partial \hat{x}}
\end{equation}

where $\partial \pmb \Lambda / \partial x = \pmb M \, \partial \pmb W / \partial x$ at the resident $\hat x$ and therefore

\begin{equation}
    \frac{\partial}{\partial \hat x}\,\bigg(\frac{\partial \pmb \Lambda}{\partial x}\bigg|_{x=\hat x}\bigg) = \pmb M \, \frac{\partial}{\partial \hat x}\,\bigg(\frac{\partial \pmb W}{\partial x}\bigg|_{x=\hat x}\bigg)
\end{equation}

and where the change in left and right eigenvectors evaluated at the resident, as the resident changes, are

\begin{equation}
    \frac{\partial \hat{\overrightarrow{v}} }{\partial \hat x} = 
    \begin{pmatrix}
        (m-1)\,\nicefrac{\partial \hat r_2}{\partial \hat x}\\
        m\,\nicefrac{\partial \hat r_2}{\partial \hat x}
    \end{pmatrix}
    \label{eq:deriv_left_eigenvector_resident}
\end{equation}

and

\begin{equation}
    \frac{\partial \hat{\overrightarrow{u}} }{\partial \hat x} = 
    \begin{pmatrix}
        (m-1)\,\nicefrac{\partial \hat r_2}{\partial \hat x}\\
        m\,\nicefrac{\partial \hat r_1}{\partial \hat x}
    \end{pmatrix}
    \label{eq:deriv_right_eigenvector_resident}
\end{equation}

where $\hat r_j(\hat x) = r_j(\hat x, \hat x)$ is the local growth rate of the resident. The local growth rate is a function of the equilibrium resource concentrations $R^*_{ij}(\hat x)$, which are themselves functions of the equilibrium population densities $N^*_j(\hat x)$ in each patch. The equilibrium population densities depend in return on the equilibrium resource concentrations $R^*_{ij}(\hat x)$ and are found, for a given resident trait value $\hat x$, by solving the resident ecological dynamics:

\begin{equation}
    \pmb \Lambda(\hat x)
    \begin{pmatrix}
        N^*_1(\hat x)\\
        N^*_2(\hat x)
    \end{pmatrix}
    =
    \begin{pmatrix}
        N^*_1(\hat x)\\
        N^*_2(\hat x) 
    \end{pmatrix}
\end{equation}

which are third-degree rational equations in $N^*_1$ and $N^*_2$, which we could not solve analytically. Therefore, we could not obtain explicit expressions for the derivatives of $N^*_1$ and $N^*_2$ with respect to the resident trait value $\hat x$, which are needed to compute the derivatives of the local growth rates $\hat r_j(\hat x)$ in the derivatives of the eigenvectors in Equations \ref{eq:deriv_left_eigenvector_resident} and \ref{eq:deriv_right_eigenvector_resident}, respectively. Instead, we used differences in selection gradient close to the singular points to calculate the convergence stability criterion.

\subsection*{Numerical procedures}

We numerically explored phenotype space for evolutionary equilibria $x^*$ by solving the root of the fitness gradient given in Equation \ref{eq:fitness_gradient_expression} across a range of parameter combinations. For each equilibrium we assessed its convergence and evolutionary stability by evaluating the fitness curvature (Equation \ref{eq:fitness_curvature_equilibrium_expression}) and the change in the direction of the gradient, respectively. Convergent stable but evolutionarily unstable equilibria were classified as branching points. The procedure is shown in the accompanying \textit{Mathematica} notebook.